---
title: "Ocean acidification and boring sponge on *Crassostrea virginica* GE"
date: "*Last run on `r format(Sys.time(), '%d %B %Y')`*"
output: 
  html_document:
    theme: simplex
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
---

---

<style>
  h2{color: #DC7633 !important}
  h1{color: #5499C7 !important}
  body{background-color: white !important}
</style>

<style>
.list-group-item.active, .list-group-item.active:hover, .list-group-item.active:focus {
    color: #212F3D;
    background-color: #EDBB99;
    font-weight: bold;}
a {
    color: #5499C7;}
.nav-tabs > li.active > a, .nav-tabs > li.active > a:hover, .nav-tabs > li.active > a:focus {
    color: #DC7633;}
body {
    font-family: "Open Sans","Helvetica Neue",Helvetica,Arial,sans-serif;
    font-size: 13px;
    line-height: 1.42857143;
    color: #212F3D;}
</style>

---

```{r setup, include = FALSE} 

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(knitr.kable.NA = '')

library("knitr")

date <- Sys.Date() # For saving with the current date
set.seed(7) # set seed


## Setting standard theme for ggplot for all plots:
theme_bove <- function(base_size = 14) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # remove the gridlines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      # remove formatting on background
      strip.background = element_blank(),
      panel.background = element_rect(fill = "transparent"),
      plot.background = element_rect(fill = "transparent", colour = NA),
      # modify legend theme
      legend.position = "none",
      legend.background = element_rect(fill = "transparent", colour = "transparent"),
      legend.key = element_rect(fill = "transparent", colour = "transparent")
    )
}


## Source the custom functions file:
#source("Code/CustomFunctions.R")

```

```{r install packages, eval = FALSE, include = FALSE}

### If any packages are not installed or need to be updated, you can look for them below:
## get gridSVG from github directly
# library(devtools) 
# devtools::install_github("cran/gridSVG")
# devtools::install_github('sinhrks/ggfortify')
# devtools::install_github("ropensci/rnaturalearthhires")

## get Bioconductor packages
if (!requireNamespace("BiocManager"))
install.packages("BiocManager")
BiocManager::install("DESeq2")

## installing WGCNA:
# source("http://bioconductor.org/biocLite.R")
BiocManager::install(c("AnnotationDbi", "impute", "GO.db", "preprocessCore"))
BiocManager::install("WGCNA", dependencies=TRUE)
BiocManager::install("arrayQualityMetrics", dependencies=TRUE) # use this arrayQualityMetrics install if using later versions of R (3.6.3 works)
#repos="http://cran.us.r-project.org"

## R version 3.6 is funky with arrayQualityMetrics so need this work around:
install.packages("ape", lib = "/Library/Frameworks/R.framework/Versions/4.2/Resources/library")
library(ape, lib.loc = "/Library/Frameworks/R.framework/Versions/4.2/Resources/library")
install.packages("magick", lib = "/Library/Frameworks/R.framework/Versions/4.2/Resources/library", dependencies = FALSE)
library(magick, lib.loc = "/Library/Frameworks/R.framework/Versions/4.2/Resources/library")
BiocManager::install("arrayQualityMetrics", type = "source", checkBuilt = TRUE, lib.loc = "/Library/Frameworks/R.framework/Versions/4.2/Resources/library")



### Used packages that need to be installed to run code (and can be sourced easier from CRAN):
needed_packages <- c("tidyverse", "dplyr", "readr", "plotly", "vegan", "data.table", "ggpubr", "pdftools", "ggrepel", "adegenet") # Specify necessary packages

not_installed <- needed_packages[!(needed_packages %in% installed.packages()[ , "Package"])] # Extract not installed packages
if(length(not_installed)) install.packages(not_installed) # Install not installed packages

```

```{r load packages, include=FALSE}

library(tidyverse)
library(dplyr)
#library(arrayQualityMetrics) # need special install above
library(ggplot2)
library(readr)
library(plotly)
library(DESeq2) # need special install above
library(vegan)
library(data.table)
library(ggpubr)
library(pdftools)
library(ggrepel)
library(adegenet)
library(ggvenn)
library(flashClust)
library(WGCNA)

source("Code/CustomFunctions.R")

#color scheme throughout
#n2800 n400 s2800 s400
color_scheme <- c("#2c7bb6", "#abd9e9", "#d7191c", "#fdae61")


```



## Expoloring data {.tabset}

```{r read in data}

# read in the counts file
counts <- read.table("Data/CVIR_featurecounts_22Jun22.txt", header = TRUE, row.names = 1)
counts <- counts[c(-1:-5, -37)] # removing columns 1-5 since we do not need them for this (and the Undetermined column)
col_names <- colnames(counts)

# Remove some of the extra stuff in the column names to match with expDesign
col_names <- gsub("X", "", col_names)
col_names <- gsub("\\_S.*", "", col_names)
colnames(counts) <- col_names

# read in the experimental design .csv
expDesign <- read.csv("Data/cvir_expDesign.csv")
expDesign <- expDesign[match(col_names, expDesign$Sample_ID),] # reorder samples to match count df
expDesign$infect <- factor(expDesign$infect)
expDesign$pCO2 <- factor(expDesign$pCO2)
expDesign$Sample_ID <- factor(expDesign$Sample_ID)
expDesign$treat <- factor(paste(expDesign$infect, expDesign$pCO2, sep = "_"))

```

<br/>

### Filtering counts

```{r unfiltered size factor plots, fig.width = 10, fig.height = 4, include = FALSE}

nrow <- nrow(counts) # number of rows/counts (38828)
countMat <- DESeqDataSetFromMatrix(counts, expDesign, ~ 1) # makes a DESeqDataSet object with count data, experimental design, and no design formula
counts_SF <- estimateSizeFactors(countMat) #  estimates the size factors using the "median ratio method" described by Equation 5 in Anders and Huber (2010)

# make dataframe of the size factors to visualize
sizeFactors <- data.frame(sample = counts_SF@colData@listData[["Sample_ID"]], treat = counts_SF@colData@listData[["treat"]], sizeFactors = counts_SF@colData@listData[["sizeFactor"]])

# plot sizeFactors
ggplot(data = sizeFactors, aes(x = sample, y = sizeFactors, fill = treat)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_manual(values=color_scheme) +
  #scale_fill_brewer(palette = 2, type = "qual") +
  geom_bar(stat = "identity") +
  ggtitle("Unfiltered sizeFactors")

```

